---
title: DATA 606 Data Project Proposal
author: Chad Bailley
---

### Data Preparation

```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
## load data
    fileLocation1 <- '02 Proficiency Data with Entity Demographics.csv'
    proficiency <-read.csv(fileLocation1, sep = ',')

## check the first few records 
    head(proficiency)

```


### Research question 

To what extent does the rate of economic disadvantage correlate with proficiency on state assessments and by how much does that correlation vary across content areas.



### Cases 

In this data set each record is a single case representing a school's rate proficiency on the state assessment for a given content area (Mathematics, English Language Arts, Science, and Social Studies) and academic year. The dataset contains multiple years and so has `r nrow(proficiency)` cases.



### Data collection 

This dataset was acquired through a request to the Michigan Department of Education.



### Type of study 

This dataset contains observational data.



### Data Source 

This data is publically available at 




### Dependent Variable

The response variable in this dataset is [PctMetProficient] which gives the percent of students that are proficient on the state assessment for the given year, building, and content area.



### Independent Variable

There are multiple independent variables in this dataset. The ones used for this project will be 
* [PctEd], a quantative which gives the percent of students that are economically disadvantaged
* [ContentAreaName], a qualitative variable giving the content area of the state assessment


### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}

```

